A decentralized network support middleware for MMOGs

### 2009-09-05 trunk/textos/PlanoDoutorado/planodoc_lugano.tex

%
% exemplo genérico de uso da classe iiufrgs.cls
% $Id: iiufrgs.tex,v 4.6 2003/11/21 22:36:13 avila Exp $
%
\documentclass[plano-doutorado]{iiufrgs}
% um tipo específico de monografia pode ser informado como parâmetro opcional:
%\documentclass[tese]{iiufrgs}
% monografias em inglês devem receber o parâmetro `english':
%\documentclass[diss,english]{iiufrgs}
% a opção `openright' pode ser usada para forçar inícios de capítulos
% em páginas ímpares
% \documentclass[openright]{iiufrgs}
% para gerar uma versão somente-frente, basta utilizar a opção `oneside':
% \documentclass[oneside]{iiufrgs}
\usepackage[T1]{fontenc}        % pacote para conj. de caracteres correto
\usepackage[latin1]{inputenc}   % pacote para acentuação
\usepackage{graphicx}           % pacote para importar figuras
\usepackage{times}              % pacote para usar fonte Adobe Times
%\usepackage{mathptmx}          % p/ usar fonte Adobe Times nas fórmulas
\usepackage{algorithm}
\usepackage{algorithmic}

@@

The \textbf{client-server architecture} meets various aspects necessary for the satisfactory implementation of games such as MMOGs, what includes a high level of control over the system as a whole, which facilitates authentication, persistence and security. But the maintenance of a traditional MMOG server is expensive, as already mentioned, besides having the server as a possible bottleneck. In order to minimize this problem some alternatives have been proposed. One of them is to use cluster computing, where a cluster, rather than a single computer, plays the role of the server. This approach has a significant increase in processing power, but does not solve all the problems of massively multiplayer online games. It should also provide the bandwidth needed to support the traffic between the server and the players.

@@

Another approach is the \textbf{peer-to-peer architecture}, where the simulation is divided among the computers involved. There can be a system without any server, where the peers, which are the machines of the players, come into some kind of agreement for the various steps of the simulation. Regarding scalability, this approach is not optimal, since ensuring such "agreement"{} is costly in terms of message exchange \cite{lamport1982bgp}. Even if one of the peers is elected to decide the course of the simulation, still persists the problem of all peers needing to exchange messages with one another. Having $n$ peers, there is a complexity of O({$n^2$}) message exchanges for each step of the simulation. Clearly, this approach is not as scalable as should be a system to support a massively multiplayer online game. Moreover, it would be necessary to distribute the storage and retrieval of the states of the game.

Some studies -- such as \cite{schiele2007rpp, rieche2007ppb, hampel2006ppa, elrhalibi2005abm, iimura2004zfg, knutsson2004pps} -- have been done to make games on peer-to-peer networks more scalable, . For example, to reduce traffic between the peers, each peer can send state updates only to whom the state changes are relevant. To achieve this goal, \cite{schiele2007rpp} suggests the division of the virtual environment into regions. The purpose of this is for each region to act as an independent set of smaller scale. Players whose avatars were in a given region could form a small peer-to-peer group and their own computers would decide the outcome of the players' actions in that area of the virtual environment. When an avatar moved from one region to another, it would leave the first group and join the group of its new region. Since the environment is considered contiguous, i.e. the avatars could move freely across adjacent regions, some of the peers in a group would keep connections to peers from groups of neighboring regions and, through these connection between adjacent groups, the necessary information for transfering a player could be obtained.

%TODO: por mais refs aqui após o ", como"
According to the work of \cite{schiele2007rpp}, each region has a coordinator. The coordinator is elected among the peers which were present and becomes responsible, only, to decide to whom each state update is relevant -- it does \emph{not} intermediate message exchanges between peers, whose communication is direct. However, this approach is based on the fact that the coordinator is reliable, which can not be guaranteed, since the software used by the player may have been modified to behave incorrectly in order to provide invalid benefits -- what is also known as \emph{cheating}. Another approach would be to elect multiple coordinators per region, but that would involve implementation of a voting mechanism, and depend on the availability of a certain number of peers to manage the region. Moreover, the communication load remains quadratic (in the worst case), since every peer must send its state to several other peers.

There are also proposals for \textbf{hybrid architectures} \cite{vilanova2008psm, chen2006pch}, using servers and peer-to-peer networks at the same time. According to these works, peers and server share the simulation of the game world. In \cite{vilanova2008psm}, the game world is divided into social space and action spaces. The first is aimed at social interactions, such as chat, exchange of in-game items, forming groups, etc., which are actions that do not impose a heavy burden on the server. However, if the players want, for example, fight -- what is the main objective in most MMOG games -- they have to request to the server to create an action space within which the interaction is faster and with little tolerance for communication delays. This space is an \emph{instance} of the game, isolated from the rest of the world, within which a limited number of players form a small peer-to-peer group, being themselves responsible for simulating the game and for updating his companions on the state of the game. To resolve inconsistencies in the simulation due to lack of message ordering, one player is chosen as a \emph{super-peer} and is responsible for ordering the events that are "heavily coupled", whose order can significantly interfere in the outcome of the game. This approach presents some problems, such as the fact that there cannot be a fight, for example, between a large number of players. Moreover, nothing guarantees that the peer chosen to be the super-peer is trustable. If it belongs to a dishonest player, it may order the events in an order such that its player is benefited. In addition, the problem of overloading the bandwidth of peers is addressed by simply setting a limit to the number of participants in each action group, when in fact it is desirable that in an MMOGs there is a lot of players interacting with each other, not only in social circumstances, but also in more dynamic situations, like fast-paced fights.

In \cite{chen2006pch}, the virtual environment is divided into regions, each managed by a peer, which acts as a sub-server. Each one of the other participants in the region sends state updates to this sub-server, which forwards the messages to the other peers in the region to which the state changes may be relevant. To provide fault tolerance, the authors suggest the use of peer clusters, consisting of the region's sub-server and backup peers. These backup peers receive the players' actions from the region manager and process them, creating a backup of the game state. In addition to providing robustness to the system in case the manager crashes, these backup nodes may also detect faults in the simulation and perform recovery procedures. However, this approach does not solve the problem of the peer elected to be the region manager not being reliable: to prevent the backup nodes from detecting an invalid simulation, the region manager may send invalid actions to them, instead of forwarding the players' real actions. Finally, if a region manager gets overloaded, it simply returns to the server part of the load assigned to it, keeping a dependency on a centralized infrastructure.

Another hybrid architecture is that of FreeMMG \cite{cecin2004fsa}. In it, the players are organized in a peer-to-peer manner in every region of the virtual environment and the server mediates the communication between different regions. Again, there is the problem of security: as the peers within a region control the simulation that occurs  there, they can also subvert it. The author proposes a probabilistic approach, using a randomly selected peer to verify the simulation in that region. It is expected that, if the peers in a given region try to subvert the game there, at least the randomly inserted node detect the invalid actions and report to the server. However, besides not completely guaranteeing security, the problem persists there may be many peers communicating with many, which may compromise the quality of the game.

@@

Finally, the authors of \cite{assiotis2006dam}, as well as several others with similar proposals \cite{ng2002msa, chertov:olb, lee2003sdl}, consider that the nodes are servers connected through a high speed and low latency network, what make their solutions only partially applicable in a scenario with highly dynamic and volatile resources, such as the peer-to-peer networks formed with volunteer resources. Such networks have some inherent problems: nodes with low availability and dependability, low bandwidth and low processing power compared to current dedicated MMOG servers maintained by MMOG developers.

@@

The aim is to investigate the use of a server system consisting of geographically distributed volunteer nodes. This system would be heterogeneous, where each node could be anything from a home computer to a cluster belonging to some company that had interest in serving the game. Some minimum processing power and communication bandwidth would be required for being part of the server, and it would be necessary to ensure somehow that each node is trusted. However, taking into account that the number of server nodes would be much smaller than the number players, it would be more feasible to ensure its reliability regarding the possibility of cheating than in a fully decentralized peer-to-peer architecture. Furthermore, the server nodes do not necessarily belong to players. They could be small local Internet service providers (ISPs), for example, who would profit by putting their resources available for a game. That would be interesting to these companies for the resources -- at least communication bandwidth -- of the ISP would be used anyway, for all the traffic from players who were also its subscribers would pass over there regardless of which server each player was connected to.

Nevertheless, as the simulation would be distributed among the server nodes, it is not necessary that all they have the same processing power or communication bandwidth owned by the huge central servers commonly used for MMOG games. To make this distribution, the game world could be divided into regions, each managed by at least one of these server nodes. More than one server node should be assigned to each region, so that the simulation would be distributed on two levels: first into regions, then within each one of them among the nodes managing it. Thus, each player would send his actions to one of these nodes, which would process them, then synchronize with other nodes in the same region and, finally, make the resulting game state persistent in a distributed way and send it to the players to whom it is relevant. Thus, it would also be provided a better availability, as each region of the virtual world would be served by a set of servers holding a synchronized game state. If any of these servers fail, one of the others could take on the players who were connected to the server that is no longer available.

@@

\section{Sincronização de estado}

Uma das questões centrais que o Cosmmus deverá tratar é a da manutenção da consistência da simulação entre os diferentes nodos servidores. Ainda que cada jogador veja apenas uma parte do mundo virtual, este mundo deve ser o mesmo para todos os servidores. Assim, se dois jogadores estão com seus avatares no mesmo lugar do ambiente virtual, eles deverão visualizar o mesmo ambiente. Outra questão importante é a ordenação dos eventos. Supondo que dois jogadores estivessem lutando no jogo, e o avatar de um deles atirasse com uma arma qualquer e o outro se movesse, a ordem desses eventos seria crucial para decidir se o outro jogador teve seu avatar morto, apenas ferido ou se evitou completamente o disparo.

No caso em que os dois jogadores estivessem conectados ao mesmo servidor, como é tradicionalmente feito por empresas de jogos, isto seria trivial, bastando verificar qual das ações chegou primeiro. Ainda que se utilizasse um mecanismo de compensação de atraso -- o \emph{timestamp} da ação de cada jogador seria igual ao tempo atual menos o seu atraso --, como é feito em Half-Life \cite{halflife}, por exemplo, não seria muito problemático ordenar estes eventos. No caso de um servidor distribuído, seria necessário que os nodos servidores entrassem em algum tipo de acordo ou sincronia no que diz respeito a essa ordenação, cuja complexidade cresceria com o número de servidores envolvidos.

Levando em conta que há várias regiões formando um mundo contíguo, e que em cada uma existe um ou mais servidores, estas questões referentes a manter a consistência da simulação se tornam ainda mais difíceis quando um avatar se aproxima de uma fronteira da região onde está localizado. Espera-se que a divisão do ambiente não seja percebida pelo jogador. Então, ao se aproximar de uma fronteira, ele deverá ser capaz de ver tudo que estiver além dela e de mover-se livremente de uma para outra. Para isto, será necessário que o servidor ao qual ele está conectado comunique-se com aquele que serve a região, além da fronteira, que está sendo visualizada pelo jogador. Serão transmitidas informações como: estado do ambiente, avatares que estejam próximos, estados desses avatares e assim por diante.

Em um trabalho do GPPD \cite{cecin2006pps}, foi proposto um esquema de sincronização de estado que consiste de dois simuladores: um conservador e um otimista. Enquanto que o otimista executa as ações dos jogadores assim que são recebidas e exibe imediatamente seu resultado na tela do jogador, o simulador conservador divide o tempo em turnos e somente executa as ações dos jogadores após haver uma sincronização. Esta sincronização consiste em que cada nodo executando a simulação receba as ações de todos os participantes envolvidos -- ou a notificação de que estes não fizeram nada --, que foram em seguida ordenadas e processadas. Somente após este passo é que o simulador conservador avança para o próximo turno. No caso da simulação otimista discrepar além do tolerável da simulação conservadora, esta sobrescreve o seu estado sobre a primeira, que continua a partir daí. No entanto, este esquema foi proposto para uma arquitetura completamente descentralizada, sendo necessário analizá-lo e fazer as adaptações necessárias para que possa ser utilizado em um sistema servidor distribuído em dois níveis, como se pretende para o Cosmmus. Existem também outros trabalhos que visam a manter consistente a simulação, como \cite{zhou2007cmh, moon2006emc, muller2006csm}, que serão também estudados e possivelmente alguns de seus princípios farão parte da solução para consistência a ser definida para o modelo final.

\section{Persistência distribuída}

No que tange ao armazenamento do estado do mundo virtual e dos jogadores, deverá haver uma maneira de fazer isto de maneira distribuída, já que não haverá um único servidor central. Poderia ser mais simples escolher um dos nodos servidores para guardar estas informações, no entanto isso poderia fazer com que o estado do jogo dependesse da disponibilidade e confiabilidade deste nodo escolhido e, ainda que estes dois requisitos fossem satisfeitos, este servidor poderia ser saturado pelo número de requisições, tornando mais lenta a recuperação dos estados nele armazenados.

O Cosmmus deverá então utilizar algum esquema de armazenamento distribuído, onde porções da informação de estado são armazenados em diferentes lugares. Além disso, deverá ser utilizada replicação para que o estado esteja disponível ainda que alguns nodos do sistema se desconectem inesperadamente. Redes lógicas baseadas em DHT, como Pastry \cite{rowstron2001psd}, Tapestry \cite{zhao2004trg}, Chord \cite{stoica2001csp} e Koorde \cite{kaashoek2003ksd}, parecem ser possíveis soluções para prover o tipo de armazenamento distribuído e redundante que se quer para o estado do jogo. Pretende-se, portanto, investigar o quanto essas redes são realmente adequadas para este fim. Existem simuladores, como por exemplo o OverSim \cite{baumgart2007ofo}, que permitem a simulação de redes overlay, tais como as que foram citadas, e através de simulação será avaliado o comportamento do sistema utilizando essas redes lógicas.

O que pode ocorrer é que seja excessivo o custo de manter uma rede como a Pastry sobre o sistema servidor, recuperando e armazenando os estados utilizando multicast no nível de aplicação e utilizando replicação, se somado ao próprio custo que os nodos servidores já enfrentarão para executar a simulação em si, recebendo ações dos jogadores e devolvendo-lhes a resposta. Para não sobrecarregar o sistema com o tráfego referente à persistência do estado do jogo, poderiam ser ajustados parâmetros, tais como freqüência de atualização e, no nível da aplicação, o nível de detalhamento do estado a ser persistido. Por exemplo, ao invés de armazenar a localização, direção e ação sendo executada por cada avatar em uma área do ambiente virtual, poderia ser armazenada apenas sua localização.

\section{Otimização do uso de largura de banda}

Pelo fato dos nodos servidores serem, idealmente, de baixo custo, é necessário economizar seus recursos tanto quanto for possível, mas sem prejudicar a qualidade do jogo. Por exemplo, podem ser feitas filtragens por interesse, onde a cada jogador só é enviado aquilo que ele pode visualizar, de forma a economizar tráfego tanto entre clientes e servidores, quanto entre os próprios servidores. Seguindo esse mesmo princípio, servidores de duas regiões diferentes podem ter uma versão antiga do estado um do outro, se não estiver havendo interação entre jogadores conectados a servidores diferentes. Dessa forma, estes servidores não precisarão atualizar-se um ao outro com tanta freqüência.

Periodicamente, ou no momento em que for necessário (e.g. dois jogadores de diferentes regiões interagindo próximo à fronteira), poderão ser trocados os estados das regiões. A granularidade dessa informação pode ser ainda mais fina, e cada servidor poderia enviar para o outro apenas os dados mais essenciais. Por exemplo, cada servidor enviaria ao outro apenas o estado do ambiente virtual próximo à fronteira entre as regiões, ao invés do estado da região inteira.

%TODO: refs.
Diversos propostas já foram realizados com o intuito de fazer filtragem por interesse \cite{bezerra2008a3, ahmed2008dai, morgan2005imm, minson2005aim}. Além de investigar qual a que melhor se aplica ao modelo proposto, e de verificar possíveis adaptações que sejam necessárias, outras maneiras de economizar os recursos do sistema servidor serão estudadas.
%dewan, eu, etc...

%TODO: mudar esse título de seção
\section{Questões filosóficas}

%A filosofia por trás do Cosmmus está na cooperação entre diversas pessoas e/ou instituições com o intuito de formar um sistema para servir um jogo online maciçamente multijogador, disponibilizando seus recursos para isso.
Apesar de, geralmente, ser necessário uma infra-estrutura central poderosa e cara \cite{feng2007wnn}, um sistema que siga a proposta aqui apresentada permitiria que recursos de empresas menores e usuários domésticos (com o mínimo exigido de capacidade de processamento e largura de banda, o que irá depender do jogo específico) fossem utilizados para formar um sistema para servir um jogo online maciçamente multijogador com qualidade comparável à dos MMOGs mais populares.

O Cosmmus tem como idéia fundamental formar um sistema servidor sobre uma rede par-a-par, ao invés de distribuir indiscriminadamente a simulação entre quaisquer jogadores, como são diversas propostas pesquisadas. Existindo essa separação, onde somente os servidores poderiam arbitrar sobre o jogo, poderiam ser formadas comunidades cujo único intuito seria o de manter jogos MMOGs funcionando, organizando quem seria autorizado a participar do sistema como um nodo servidor e assim por diante. Poderia haver um esquema semelhante ao utilizado em diversos projetos de software livre, na Wikipédia \cite{adler2007cdr} e outros, onde o indivíduo que teve a idéia inicial vai gradualmente delegando autoridade a membros da comunidade, de acordo com o quanto cada um colaborou, criando uma hierarquia, onde há diferentes níveis de permissão, mas todos participam em um esforço comum.

Além dessa questão filosófica, para um servidor de MMOGs, que age como árbitro central e por onde toda a simulação tem que passar, é mais fácil detectar e impedir trapaça do que em um MMOG com a simulação completamente descentralizada, onde o computador de cada jogador decide em parte o que irá acontecer como resultado das ações no jogo. No entanto, mesmo em uma rede colaborativa como a que se tem em mente nesta proposta, não há garantias de que alguns dos integrantes não irão tentar subverter as regras do jogos, obtendo vantagens de forma injusta. Porém, imagina-se um sistema de reputação. Ao ingressar no sistema servidor, cada participante terá um registro, armazenado em algum local confiável, de como tem realizado a simulação do jogo.

A princípio, seria designada ao novo servidor uma porção do mundo que seja considerada de importância mínima, de tal modo que tentativas de subversão por parte dele causarão um dano muito pequeno ao jogo, se causarem. Tanto jogadores, quanto outros nodos servidores, poderão reportar casos como esse, de forma a punir ou banir o nodo que agiu de maneira maliciosa. Integrantes honestos do sistema servidor irão ganhando pontos de reputação com o passar do tempo, podendo administrar porções e regiões do mundo do jogo cada vez mais importantes. Acredita-se que seja desmotivadora o suficiente a idéia de destruir a reputação conseguida após um longo tempo de colaboração para o jogo por causa de uma tentativa de trapaça.

Dessa maneira, diversos MMOGs poderiam surgir e se popularizar com mais facilidade, abrindo campo para empresas pequenas e grupos independentes de desenvolvimento de jogos. A cada jogo estaria associada uma comunidade, que se auto-regularia. O nível de justiça ou trapaça em cada um deles iria depender de quem fossem os responsáveis. Comunidades bem organizadas poderiam fazer com que um jogo fosse tão justo quanto um administrado por provedores com contratos comerciais assinados, por exemplo.

>>>>

%
% exemplo genérico de uso da classe iiufrgs.cls
% $Id: iiufrgs.tex,v 4.6 2003/11/21 22:36:13 avila Exp $
%
\documentclass[plano-doutorado]{iiufrgs}
% um tipo específico de monografia pode ser informado como parâmetro opcional:
%\documentclass[tese]{iiufrgs}
% monografias em inglês devem receber o parâmetro `english':
%\documentclass[diss,english]{iiufrgs}
% a opção `openright' pode ser usada para forçar inícios de capítulos
% em páginas ímpares
% \documentclass[openright]{iiufrgs}
% para gerar uma versão somente-frente, basta utilizar a opção `oneside':
% \documentclass[oneside]{iiufrgs}
\usepackage[T1]{fontenc}        % pacote para conj. de caracteres correto
\usepackage[latin1]{inputenc}   % pacote para acentuação
\usepackage{graphicx}           % pacote para importar figuras
\usepackage{times}              % pacote para usar fonte Adobe Times
%\usepackage{mathptmx}          % p/ usar fonte Adobe Times nas fórmulas
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[english]{babel}

@@

The \textbf{client-server architecture} meets various aspects necessary for the satisfactory implementation of games such as MMOGs, what includes a high level of control over the system as a whole, which facilitates authentication, persistence and security. But the maintenance of a traditional MMOG server is expensive, as already mentioned, besides making the server a possible bottleneck. In order to minimize this problem some alternatives have been proposed. One of them is to use cluster computing, where a cluster, rather than a single computer, plays the role of the server. This approach has a significant increase in processing power, but does not solve all the problems of massively multiplayer online games. It should also provide the bandwidth needed to support the traffic between the server and the players.

@@

Another approach is the \textbf{peer-to-peer architecture}, where the simulation is divided among the computers involved. There can be a system without any server, where the peers, which are the machines of the players, come into some kind of agreement for the various steps of the simulation, calculating the game state resulting of each action of the participants. Regarding scalability, this approach is not optimal, since ensuring such "agreement"{} is costly in terms of message exchange \cite{lamport1982bgp}. Even if one of the peers is elected to decide the course of the simulation, still persists the problem of all peers potentially needing to exchange messages with one another. Having $n$ peers, there is a complexity of O({$n^2$}) message exchanges for each step of the simulation, in the worst case. Clearly, this approach is not as scalable as should be a system to support a massively multiplayer online game. Moreover, it would be necessary to distribute the storage and retrieval of the states of the game.

Some studies -- such as \cite{schiele2007rpp, rieche2007ppb, hampel2006ppa, elrhalibi2005abm, iimura2004zfg, knutsson2004pps} -- have been done to make games on peer-to-peer networks more scalable. For example, to reduce traffic between the peers, each peer can send state updates only to whom the state changes are relevant. To achieve this goal, \cite{schiele2007rpp} suggests the division of the virtual environment into regions. The purpose of this is for each region to act as an independent set of smaller scale. Players whose avatars were in a given region could form a small peer-to-peer group and their own computers would decide the outcome of the players' actions in that area of the virtual environment. When an avatar moved from one region to another, it would leave the first group and join the group of its new region. Since the environment is considered contiguous, i.e. the avatars could move freely across adjacent regions, some of the peers in a group would keep connections to peers from groups of neighboring regions and, through these connections between adjacent groups, the necessary information for transferring a player could be obtained.

%TODO: por mais refs aqui após o ", como"
Still according to the work of \cite{schiele2007rpp}, each region has a coordinator. The coordinator is elected among the peers which were present in the region and becomes responsible, only, to decide to whom each state update is relevant -- it does \emph{not} intermediate message exchanges between peers, whose communication is direct. However, this approach is based on the fact that the coordinator is reliable, which can not be guaranteed, since the software used by the player may have been modified to behave incorrectly in order to provide invalid benefits -- what is also known as \emph{cheating}. Another approach would be to elect multiple coordinators per region, but that would involve the implementation of a voting mechanism, and depend on the availability of a certain number of peers to manage the region. Moreover, the communication load remains quadratic (in the worst case), since every peer must send its state to several other peers.

There are also proposals for \textbf{hybrid architectures} \cite{vilanova2008psm, chen2006pch}, using servers and peer-to-peer networks at the same time. According to these works, peers and server share the simulation of the game world. In \cite{vilanova2008psm}, the game world is divided into social space and action spaces. The first is aimed at social interactions, such as chat, exchange of in-game items, forming groups, etc., which are actions that do not impose a heavy burden on the server. However, if the players want, for example, fight -- what is the main objective in most MMOGs -- they have to request to the server the creation of an action space within which the interaction is faster and with little tolerance for communication delays. This space is an \emph{instance} of the game, isolated from the rest of the game world, within which a limited number of players form a small peer-to-peer group, being themselves responsible for performing the simulation and for updating his companions on the game state. To resolve inconsistencies in the simulation due to lack of message ordering, one player is chosen as a \emph{super-peer} and is responsible for ordering the events that are "heavily coupled", whose order can significantly interfere in the outcome of the game. This approach presents some problems, such as the fact that there cannot be a fight, for example, between a large number of players. Moreover, nothing guarantees that the peer chosen to be the super-peer is trustable. If it belongs to a dishonest player, it may put the events in an order that provides an invalid advantage over the other players. In addition, the problem of overloading the bandwidth of peers is addressed by simply setting a limit to the number of participants in each action group, when in fact it is desirable that in an MMOGs there is a lot of players interacting with each other, not only in social circumstances, but also in more dynamic situations, like fast-paced fights.

In \cite{chen2006pch}, the virtual environment is divided into regions, each managed by a peer, which acts as a sub-server. Each one of the other participants in the region sends state updates to this sub-server, which forwards the messages to the other peers in the region to which the state changes may be relevant. To provide fault tolerance, the authors suggest the use of peer clusters, consisting of the region's sub-server and backup peers. These backup peers receive the players' actions from the region manager and process them, creating a backup of the game state. In addition to providing robustness to the system in case the manager crashes, these backup nodes may also detect faults in the simulation and perform recovery procedures. However, this approach does not solve the problem of the peer elected to be the region manager 	not being reliable: to prevent the backup nodes from detecting an invalid simulation, the region manager may send invalid actions to them, instead of forwarding the players' real actions. Finally, if a region manager gets overloaded, it may simply return to the server part of the load assigned to it, keeping a dependence on a centralized infrastructure.

Another hybrid architecture is that of FreeMMG \cite{cecin2004fsa}. In it, the players are organized in a peer-to-peer manner in every region of the virtual environment and the server mediates the communication between different regions. Again, there is the problem of security: as the peers within a region control the simulation that occurs  there, they can also subvert it. The author proposes a probabilistic approach, using a randomly selected peer to verify the simulation in that region. It is expected that, if the peers in a given region try to subvert the game there, at least the randomly inserted node detects the invalid actions and reports to the server. However, besides not completely guaranteeing security, persists the problem that there may be many peers communicating with many, which may compromise the quality of the game.

@@

Finally, the authors of \cite{assiotis2006dam}, as well as several others with similar proposals \cite{ng2002msa, chertov:olb, lee2003sdl}, consider that the nodes are servers connected through a high speed and low latency network, what makes their solutions only partially applicable in a scenario with highly dynamic and volatile resources, such as the peer-to-peer networks formed with volunteer resources. Such networks have some inherent problems: nodes with low availability and dependability, low bandwidth and low processing power compared to current dedicated MMOG servers maintained by large game companies.

@@

This doctorate plan considers the use of a server system consisting of geographically distributed volunteer nodes. This system would be heterogeneous, where each node could be anything from a home computer to a cluster belonging to some company that had interest in serving the game. Some minimum processing power and communication bandwidth would be required for being part of the server, and it would be necessary to ensure somehow that each node is trusted. However, taking into account that the number of server nodes would be much smaller than the number of players, it would be more feasible to ensure its reliability regarding the possibility of cheating than in a fully decentralized peer-to-peer architecture. Furthermore, the server nodes do not necessarily belong to players. They could be small local Internet service providers (ISPs), for example, who would profit by putting their resources available for a game. That would be interesting to these companies for the resources -- at least communication bandwidth -- of the ISP would be used anyway, for all the traffic from players who were also its subscribers would pass over there regardless of which server each player was connected to.

Nevertheless, as the simulation would be distributed among the server nodes, it is not necessary that all they have the same processing power or communication bandwidth owned by the huge central servers commonly used for MMOGs. To perform this distribution, the game world could be divided into regions, each managed by at least one of these server nodes. More than one server node should be assigned to each region, so that the simulation would be distributed on two levels: first into regions, then within each one of them among the nodes managing it. Thus, each player would send his actions to one of these nodes, which would process them, then synchronize with other nodes in the same region and, finally, make the resulting game state persistent in a distributed way and send it to the players to whom it is relevant. Thus, it would also be provided a better availability, as each region of the virtual world would be served by a set of servers holding a synchronized game state. If any of these servers fail, one of the others could take on the players who were connected to the server that is no longer available.

@@

\section{State synchronization}

One of the key issues that will be handled by Cosmmus is maintaining the consistency of the simulation among the different server nodes. Even if each player sees just a part of the virtual world, this world must be the same for all servers. Thus, if two players are with their avatars in the same place of the virtual environment, they should see the same scene. Another important issue is the ordering of events. Assuming that two players are fighting in the game, the avatar of one of them shoots with a gun and the other avatar moves, the order of these events would be crucial to decide whether the latter player had his avatar dead, only wounded or completely avoided the bullet.

In the event that both players are connected to the same server, as it is traditionally done by game developing companies, this task would be trivial, by simply checking which actions come first. Even if a mechanism for delay compensation -- the \emph{timestamp} of the action of each player would be equal to the current time minus the player's connection latency -- as it is done in Half-Life \cite{halflife}, for example, it would not be too problematic to sort these events. In the case of a distributed server, it would require the server nodes to enter into some kind of agreement or synchronization regarding this event ordering, whose complexity would grow with the increase of the number of servers involved.

Taking into consideration that there are several regions forming a contiguous world, and that in each one of them there is one or more servers, these issues to maintain the consistency of the simulation becomes even more difficult. It is expected that the division of the environment is seamless to the player. Then, when approaching a region boundary, the player should be able to see whatever is beyond it and to move freely across the border. For this, it will be necessary that the server to which he is connected to communicate with the one that serves the region beyond the border which is being viewed by the player. It will provide information such as the state of the environment and of avatars who are nearby and of other objects in the game.

In \cite{cecin2006pps}, it was proposed a state synchronization scheme that consists of two simulators: one conservative and one optimistic. The optimistic one processes each player action immediately after it is received and its resulting state becomes available to the player and something is show on his screen. The conservative simulator, however, divides the time in turns and only performs the actions of the players after performing a synchronization. At each turn, it waits for the arrival of the actions of all players involved -- or some explicit notification that they did nothing -- which ared then sorted and processed. Only after this step the conservative simulator advances to the next turn. In the case of the optimistic simulation deviate beyond a tolerable limit from the conservative one, the latter overwrites its state on the first, which continues from there. However, this scheme was proposed for a fully decentralized architecture, being necessary to analyze it and make necessary adjustments so it can be used in a server system distributed in two levels, as intended for Cosmmus. There are also other works that aim to maintain a consistent simulation, such as \cite{zhou2007cmh, moon2006emc, muller2006csm}, which will also be studied and possibly some of its principles will be part of the solution for consistency designed for the final model.

\section{Distributed persistence}

Regarding the storage of the state of the virtual world and the players, it will be devised a way to do this in a distributed manner, since there will not be an unique central server. It might be simpler to choose one of the server nodes to save this information. However, this could cause the game state to depended on the availability and reliability of the chosen node and, even if these two conditions were satisfied, the node's connection could be easily saturated by the great number of requests, slowing down the recovery of the states stored in it.

Cosmmus should then use a distributed storage scheme, where portions of the state information are stored in different places. Furthermore, it should replication, so that the state becomes available even if some nodes of the system disconnect or crash unexpectedly. DHT-based overlay networks, such as Pastry \cite{rowstron2001psd}, Tapestry \cite{zhao2004trg}, Chord \cite{stoica2001csp} and Koord \cite{kaashoek2003ksd} seem to be possible solutions to provide the kind of distributed and redundant storage desired. It is intended, therefore, to investigate how suitable these networks really are for this purpose. There are simulators, such as OverSim \cite{baumgart2007ofo}, allowing the simulation of overlay networks, such as those mentioned, and through simulation it be will assessed the behavior of the system when using some of these logical networks.

A problem, however, is that the computational cost of using those DHT-based networks may become too high, for it would be necessary to maintain the structure of a network such as Pastry on the server system, retrieving and storing the game state with application level multicast and replication, added to the own cost that the server nodes handle to perform the game world simulation itself, receiving players' actions and returning them a resulting state. To avoid overloading the system with traffic related only to persistence of the game state, some parameters could be adjusted, like the state update frequency and, at the application level, the level of detail of the state to be persisted. For example, instead of storing the location, direction and action being performed by each avatar in an area of the virtual environment, it could be stored only its location.

\section{Bandwidth usage optimization}

As the server nodes are, ideally, of low cost, it is necessary to save their resources as much as possible, but without harming the game quality. For example, it can be made some interest-based filtering, being sent to each player only what he can see, in order to save as much bandwidth between clients and servers, and between the servers. Following this same principle, servers from two different regions may have an old version of the state of each other, if there is no interaction between players whose avatars are in different regions. Thus, these servers do not need to update each other so often.

Periodically, or when it is needed (e.g. two players from different regions interacting near the border), the states of the regions may be exchanged between the servers managing them. The granularity of this information can be even thinner, and each server could send to the other only the most essential data. For example, each server would send to the other only the state of the virtual environment near the border between the regions, rather than the state of the entire region.

Several works have been made in order to make filtering interest-based message filtering \cite{bezerra2008a3, ahmed2008dai, morgan2005imm, minson2005aim}. In addition to studying which one best fits the proposed model, and evaluate the possible adjustments that are necessary, other ways to save the resources of the server system will be investigated.

\section{High-level application programming interface}

In order to offer to application (game) developers an easy-to-use set of services, a high-level application programming interface must be created for Cosmmus. This set of services will compose the Cosmmus middleware, which will encapsulate all the low level operations regarding the distribution of the game. There are some options to define this distribution model -- and, also, the abstraction provided to the game developer --, like distributed shared memory, publish/subscribe, message passing and others. Each one of these presents both advantages and disadvantages, but the publish/subscribe seems to be the most suited for the development of MMOGs, since the entities (avatars and objects) present in the game's virtual world imply state update messages, in which every player may be interested or not.

%TODO: mudar esse título de seção
\section{Philosophical questions}

%A filosofia por trás do Cosmmus está na cooperação entre diversas pessoas e/ou instituições com o intuito de formar um sistema para servir um jogo online maciçamente multijogador, disponibilizando seus recursos para isso.

Although generally it is needed a powerful, centralized and expensive \cite{feng2007wnn} infrastructure to support an MMOG, a system that follows the approach outlined here would allow the resources of smaller businesses and home users (with the required minimum of processing power and connection bandwidth, which will depend on the particular game) to be used in order to form a system to serve a massively multiplayer online game with quality comparable to the most popular MMOGs.

The fundamental idea behind Cosmmus is to form a server system on a peer-to-peer network, rather than indiscriminately distributing the simulation among all players, as are several proposals surveyed. With such separation, where only the server nodes could decide the progress of the game, it could be formed communities whose sole purpose would be to keep MMOGs up and running, organizing who would be allowed to participate in the system as a server node, with which permissions, and so on. There could be a scheme similar to that used in various free software projects, in Wikipedia \cite{adler2007cdr} and others, where the individual -- or company -- who had the initial idea gradually delegates authority to community members, according to how much each one cooperated, creating a hierarchy, with different levels of permission, but with all participating in a common effort.

In addition to this philosophical question, to an MMOG server, which acts as referee and where the actual simulation is performed, it is easier to detect and prevent cheating than in an MMOG with a completely decentralized simulation, where each player's computer decides in part what will happen as a result of players' actions in the game. However, even in a collaborative network as the one idealized in this study plan, there is no guarantee that some of the members will not try to subvert the rules of the game, providing unfair advantages to some players, or even just destroying part of the game state. However, a reputation system is imagined. When entering the server system, each participant will have a record, stored in a trusted location, of how he has performed the simulation of the game.

This reputation system would work as follows: at first, it would be assigned to the new server a portion of the world which is considered of minimal importance, so that attempts to subvert that part of the game environment will cause very little damage to the game, if any. Both players and servers may report such cases in order to punish or even banish the node that acted maliciously. Honest members of the server system will gain reputation points with time, and they will manage increasingly important portions of the game world. It is believed to be enough demotivating the idea of destroying a good reputation, gained after a long period of honest collaboration with the game because of an cheating attempt.

This way, many MMOGs could arise and become popular more easily, opening this market to small companies and independent game developers. Every game would be associated with a self-regulating community. The level of justice or tampering of each one of them would depend on who were responsible. A well organized community could make a game as fair as those managed by large companies with expensive and powerful central servers, for example.
